Switch equivalence and weak prime domains for fusions - quotients

### 2024-09-01 quozienti.tex
minori

\newcommand{\cmd}{\mathbf{IC}}

\newcommand{\cori}{\mathbf{CR}}

@@

@@

\begin{definition}
	Let $\X$ be a category, a subcategory $\Y$ of $\X$ is \emph{coreflective} if the inclusion functor $I_\Y\colon \Y\to \X$ has a right adjoint $R_\Y$. A \emph{coreflective pair} $(\Y, R_\Y)$ is a pair made by a coreflective subcategory $\Y$ of $\X$ and a chosen right adjoint to $I_\Y$. A morphism between two coreflective pairs $(\Y, R_\Y)$ and $(\Z, R_\Z)$ is a functor $F\colon \Y \to \Z$ making the following diagrams commutative.
	\[\xymatrix{\Y  \ar[dr]_{I_\Y}  \ar[rr]^{F} &&\Z  \ar[dl]^{I_\Z}&&\X \ar[dr]^{R_\Z} \ar[dl]_{R_\Y} \\ & \X && \Y \ar[rr]_{F} && \Z }\]

@@

@@

Associated to a coreflective subcategory $\Y$ of $\X$ we have a class of arrows which enjoys some desiderable properties.

In general, the class $\mathcal{R}_{(\Y, R_\Y)}$ is not closed under pushout. Nonetheless such property is desirable and often enjoyed in the setting of DPO-rewriting. We are thus lead to the following definition.

 \begin{proposition}\label{prop:funct}
 	contenuto...
 \end{proposition}
 \begin{proof}
 	contenuto...
 \end{proof}

@@

\subsection{A left adjoint to $\mathcal{R}_{(-)}$}

\begin{setting}Let $\X$ be a category with a proper factorization system $(\mathcal{E}, \mathcal{M})$. Our abstract setting to deal with fusions begins with a coreflective subcategory $I\colon \Y\to \X$ with coreflector $F\colon \X\to \Y$ with the following properties:
	\begin{enumerate}
		\item the counit $\epsilon \colon I\circ F\to \id{\X}$ has all components in $\mathcal{M}$; 
		\item given a pushout as the one below, if $F(f)$ is mono then $F(h)$ is mono too.
		\[\xymatrix{X \ar[r]^{g} \ar[d]_{f}& Z \ar[d]^{h}\\ Y\ar[r]_{k} & W}\]
	\end{enumerate}

@@

\begin{lemma}	
Let $\X$ be an $\mathcal{M}$-adhesive category $\X$  with a proper and stable factorization system $(\mathcal{E}, \mathcal{M})$. Suppose that $\mathcal{M}$ has $\mathcal{M}$-unions. Then there exists a corefelctive subcategory $I\colon \Y\to \X$ satisfying the assumptions of our setting.
\end{lemma}
\begin{proof}
	contenuto...
\end{proof}

>>>>

\newcommand{\cori}{\mathbf{CS}}
\newcommand{\cla}{\mathbf{CL}}
\newcommand{\pos}{\mathbf{PO}}
\newcommand{\gori}{\mathbf{GS}}

@@

\newcommand{\dm}[1]{\mathsf{dom}({#1})}

@@

\begin{definition}
	Let $\X$ be a category, a full subcategory $\Y$ of $\X$ is \emph{coreflective} if the inclusion functor $I_\Y\colon \Y\to \X$ has a right adjoint $R_\Y$, called \emph{coreflector}.  A morphism between two coreflective subcategories $\Y$ and $\Z$ is a functor $F\colon \Y \to \Z$ such that $I_\Z \circ F=I_\Y$ and $F\circ R_\Y$ is isomorphic to $R_\Z$.

@@

\begin{remark}
	Notice that, since adjoints are unique up to isomorphism, being a morphism $\Y\to \Z$ of $\cori(\X)$ does not depends on the choice of coreflectors for $\Y$ and $\Z$.
\end{remark}

\begin{remark}\label{rem:ff}
	Let $F\colon \Y \to \Z$ be a morphism in $\cori(\X)$. Since $I_\Z\circ F=I_\Y$, then $\Y$ must be a full subcategory of $\Z$ and $F$ must coincide with its inclusion. In particular, $F$ is full and faithful and $\cori(\X)$ is a ordered class.
\end{remark}

@@

Associated to a coreflective subcategory $\Y$ of $\X$ we have a class of arrows which enjoys some useful properties.

 \begin{definition}
 	Let $\Y$ be a coreflective subcategory of $\X$ with coreflector $R_\Y$, we define the class $\mathcal{R}_\Y$ as the class of all arrows in $\X$ whose image through $R_\Y$ is a mono in $\Y$.
 \end{definition}

 \begin{remark}
 	Let $I_\Y\colon \Y \to \X$ be the inclusion functor of a coreflective subcategory $\Y$. If $R_\Y$ and $R'_\Y$ are coreflectors, then we have a natural isomorphism $\phi\colon R_\Y\to R'_\Y$ and so, for every $f\colon X\to Y$ there is a square
 	\[\xymatrix{ \ar[d]_{\phi_X} R_\Y(X) \ar[r]^{R_\Y(f)}& R_\Y(Y) \ar[d]^{\phi_Y}\\R'_\Y(X) \ar[r]_{R_\Y(f)} & R'_\Y(Y)}\]

 In particular, $R_\Y(f)$ is mono if and only if $R'_\Y(f)$ is so, thus the definition of $\mathcal{R}_\Y$ is independent from the choice of coreflector.
 \end{remark}

 \begin{proposition}
Let $\Y$ be a coreflective subcategory of $\X$. Then the following hold true:
\begin{enumerate}
	\item $\mathcal{R}_\Y$ contains all monos;
	\item $\mathcal{R}_\Y$ is stable under pullbacks;
	\item given $f\colon X\to Y$ and $g\colon Y\to Z$, if $g\circ f$ is in $\mathcal{R}_\Y$ then $f$ is in $\mathcal{R}_\Y$ too;
	\item given $f\colon X\to Y$ in $\mathcal{R}_\Y$ and $g,h\colon I_\Y(Z)\rightrightarrows X$, if $f\circ g = f\circ h$, then $g=h$.
\end{enumerate}
 \end{proposition}
  \begin{proof}\begin{enumerate}
  		\item This follows at once from the fact that $R_\Y$, being a right adjoint, preserves monos.
  		\item Consider the following two diagrams, with $f\in \mathcal{R}_\Y$, in which the left one is a pullback. $R_\Y$ preserves limits, thus the right square is a pullback too.
  		\[\xymatrix@C=30pt{P \ar[r]^{p_1} \ar[d]_{p_2}& X \ar[d]^{f} & R_\Y(P) \ar[r]^{R_\Y(p_1)} \ar[d]_{R_\Y(p_2)} & R_\Y(X) \ar[d]^{R_\Y(f)}\\Z \ar[r]_{g}& Y & R_\Y(Z) \ar[r]_{R_\Y(g)} & R_\Y(Y)}\]
  		By definition, $R_\Y(f)$ is a mono in $\Y$, but this implies that $R_\Y(g)$ is a mono too, proving $g\in \mathcal{R}_\Y$.
  		\item We know that $R_\Y(g\circ f)$ is mono, thus $R_\Y(f)$ is mono too.
  		\item 	Since $f\circ g=f\circ h$ and $f$ is in $\mathcal{R}_\Y$ we can conclude that $R_\Y(g)=R_\Y(h)$. Thus we have:
  		\begin{align*}
  			g\circ \epsilon_{I_\Y(Z)}&=\epsilon_X\circ I_\Y(R_\Y(g))\\&=\epsilon_X\circ I_\Y(R_\Y(h))\\&=h\circ \epsilon_{I_\Y(Z)}
  		\end{align*}
  		By \Cref{cor:counit},  $\epsilon_{I_\Y(Z)}$ is an isomorphism and so $g=h$. \qedhere
  	\end{enumerate}
  \end{proof}

In general, the class $\mathcal{R}_{\Y}$ is not stable under pushout. Nonetheless such property is desirable. We are thus lead to the following definition.

 \begin{definition}
A coreflective subcategory is \emph{good} if  $\mathcal{R}_\Y$ is stable under pushout. $\gori(\X)$ will denote the subcategory of $\cori(\X)$ given by good coreflective subcategories. 
 \end{definition}

\iffalse 
 \begin{remark}\label{rem:good}
Let $\Y$ be a full subcategory of $\X$ with coreflector $R_\Y$.  Let also $f\colon X\to Y$ be in $\mathcal{R}_\Y$ and consider the following two diagrams, where the left one is a pushout. Then $\Y$ is a good subcategory if and only if $R_{\Y}()$
 \end{remark}
 \fi

 We end this section extending the construction of $\mathcal{R}_\Y$ to a functor on $\cori(\X)$.

 \begin{proposition}\label{prop:funct}
 Let $F\colon \Y \to \Z$ be a morphism of $\cori(\X)$, then $\mathcal{R}_\Z \subseteq \mathcal{R}_{\Y}$.
 \end{proposition}
 \begin{proof}
 Let $f$ be an arrow in $\mathcal{R}_\Z$. Since $F$ is an arrow in $\cori(\X)$ we know that there exists a natural isomorphism $\phi\colon F\circ R_\Y\to R_\Z$, therefore $F(R_\Y(f))$ is a monomorphism. By \Cref{rem:ff} we know that $F$ is full and faithful and so reflects mono, implying that $f$ belongs to $\mathcal{R}_\Y$ too.
 \end{proof}

 \begin{corollary}\label{cor:funct}
 Let $\cla(\X)$ be the category associated to the collection of all classes of a category $\X$, ordered by the inclusion. Then there exists a functor $\mathcal{R}_{(-)}\colon \cori(\X)^{op}\to \cla(\X) $.

 Let, moreover, $\pos(\X),$ be the subcategory of $\cla(\X)$  given by classes of arrows stable under pushouts, Then  $\mathcal{R}^G_{(-)}\colon \gori(\X)^{op}\to \pos(\X) $ fitting in the rectangle below, where the vertical arrows are inclusions.
 \[\xymatrix{\cori(\X)^{op} \ar[r]^{\mathcal{R}_{(-)}}& \cla(\X)\\ \gori(\X)^{op} \ar[u]^{J} \ar[r]_{\mathcal{R}^{G}_{(-)}}& \pos(\X) \ar[u]_{I}}\]
 \end{corollary}

@@

\subsection{Left adjoints to $\mathcal{R}_{(-)}$ and $\mathcal{R}^G_{(-)}$}

In this section we will show that, under suitable hypotheses, the functors $\mathcal{R}_{(-)}$ and $\mathcal{R}^G_{(-)}$ built in the previous sections, both have left adjoints. To do so, we need some preliminary work to be done.

\begin{definition}
	Let $\mathcal{M}$ be a class of arrows in a category $\X$ and let $\{m_i\}_{i\in I}$ be a family of objects in $\mathcal{M}/X$. We say that $\{m_i\}_{i\in I}$ has the \emph{epic $\mathcal{M}$-sum property} if:
	\begin{enumerate}
		\item it has a  coproduct $m\colon M\to X$;
		\item the family  of coprojection $\{k_i\}_{i\in I}$ is jointly epic in $\X$.
	\end{enumerate}

	We say that $\X$ has \emph{epic $\mathcal{M}$-sums} if every family in $\sub{M}{X}{X}$ has the epic $\mathcal{M}$-sum property.
\end{definition}

\begin{remark}
	contenuto...
\end{remark}
\begin{example}\todo{controesempio spazi topologici}
\end{example}
\begin{example}
\todo{insiemi}
\end{example}

\begin{example}
	\todo{adesive finitare}
\end{example}
\begin{example}
	\todo{simple graph}
\end{example}

We are now ready to construct a left adjoint to $\mathcal{R}_{(-)}$.

\begin{definition} 
	Let $\X$ be a category and $\mathcal{A}$ a class of arrows of it. We define the category $\C(\mathcal{A})$ as the full subcategory of $\X$ given by the objects $Z\in \X$ such that, for every  pair of arrows $g,h\colon Z\rightrightarrows X$, if $f\circ g=f\circ h$ for some $f\colon X\to Y$ belonging to $\mathcal{A}$, then $g=h$.
\end{definition}

A first thing that we can notice about $\C(\mathcal{A})$ is that, in the presence of a factorization system, is its closure under epimorphisms, as shown by the following proposition.

\begin{proposition}\label{prop:ima}
 $\mathcal{A}$ be a class of arrows in a category $\X$ If $e\colon X\to Y$ is an epimorphism with domain in $\C(\mathcal{A})$, then $Y$ belongs to $\C(\mathcal{A})$ too.
\end{proposition}
\begin{proof}
	Let $f\colon Z\to V$ be an arrow of $\mathcal{A}$ and take two other arrows $g,h\colon Y\to Z$ such that $f\circ g=f\circ h$. Then we also know that
	\[f\circ g\circ e= f\circ h\circ e\]
	By hypothesis $X$ is in $\C(\mathcal{A})$, so that $g\circ e=f\circ e$. The thesis now follows since $e$ is an epimorphism.
\end{proof}
\begin{example}
	\todo{spazi topologici puntati - controesempio}
\end{example}

\begin{lemma}
Let $\X$ be a category with a proper factorization system $(\mathcal{E}, \mathcal{M})$. Suppose, moreover, that $\X$ has the epic $\mathcal{M}$-sum property.  Then for every $\mathcal{A}\in \cla(\X)$, $\C(\mathcal{A})$ is a coreflective subcategory of $\X$.
\end{lemma}

\begin{proof}
For every object $X$ of $\X$, let us define
\[\mathcal{F}_X:=\{m\in \mathcal{M}/X \mid \dm{m} \text{ belongs to } \C(\mathcal{A})\}\] 
We can then define $R(X)$ as the domain of the sum over $\mathcal{F}_X$ in $\mathcal{M}/X$. In particular, we have an arrow $\epsilon_X\colon R(X)\to X$, belonging to $\mathcal{M}$.

First of all, we have to prove that $R(X)$ so defined is an object of $\C(\mathcal{A})$. Let thus $f\colon A\to Y$ be an arrow in $\mathcal{A}$ and suppose that there are $g,h\colon R(X)\to A$ such that  $f\circ h= f \circ g$. Now, for every $m\colon M\to X$ in $\mathcal{F}_X$ we have a coprojection $k_m\colon m\to \epsilon_X$. Precomposing with it we get
\[f\circ h \circ k_m=f\circ g\circ k_m\]
By definition $M$ is in $\C(\mathcal{A})$, so we can conclude that $h\circ k_m=g\circ k_m$. Since  this holds for every $m\in \mathcal{F}_Z$, the epic $\mathcal{M}$-sum property of $\X$ entails $f=g$ as wanted.

To conclude, is it now enough to show that $\epsilon_X$ has the universal property of a counit. Let $Y$ be an object of $\C(\mathcal{A})$ with an arrow $f\colon Y\to X$. We can factor $f$ it as $m\circ e$ with $e\colon Y\to E$ and $m\colon E\to X$ in $\mathcal{M}$. By \Cref{prop:ima} we know that $E$ is in $\C(\mathcal{A})$, thus there exists a coprojection $k_m\colon m\to \epsilon_X$. 
\end{proof}

\begin{example}
	\todo{spazi topologici puntati - controesempio 2}
\end{example}

\begin{corollary}
	contenuto...
\end{corollary}
\iffalse 
\begin{setting}Let $\X$ be a category with a proper factorization system $(\mathcal{E}, \mathcal{M})$. Our abstract setting to deal with fusions begins with a coreflective subcategory $I\colon \Y\to \X$ with coreflector $F\colon \X\to \Y$ with the following properties:
	\begin{enumerate}
		\item the counit $\epsilon \colon I\circ F\to \id{\X}$ has all components in $\mathcal{M}$; 
		\item given a pushout as the one below, if $F(f)$ is mono then $F(h)$ is mono too.
		\[\xymatrix{X \ar[r]^{g} \ar[d]_{f}& Z \ar[d]^{h}\\ Y\ar[r]_{k} & W}\]
	\end{enumerate}

@@

\begin{lemma}	
Let $\X$ be an $\mathcal{M}$-adhesive category $\X$  with a proper and stable factorization system $(\mathcal{E}, \mathcal{M})$. Suppose that $\mathcal{M}$ has $\mathcal{M}$-unions. Then there exists a corefelctive subcategory $I\colon \Y\to \X$ satisfying the assumptions of our setting.
\end{lemma}
\begin{proof}
	contenuto...
\end{proof}
\fi
